---
title: "DATA 205 Capstone"
subtitle: "Food inspections and closed establishments"
author: "Ousmane Kandji"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(knitr)
library(lubridate)
```

## Project idea

This report uses:
- **FoodInspectionData.csv** for all inspection outcomes
- **Closed_Food_Establishments.csv** for inspections that ended in **Facility Closed**

Goal: keep a simple public-health story with 3 visuals:
1) overall inspection outcomes,
2) most common issues among closed establishments,
3) monthly trend in closures.

```{r load-food}
food <- read.csv("FoodInspectionData.csv", stringsAsFactors = FALSE) %>%
  distinct() %>%
  mutate(
    Inspection.results = str_trim(as.character(Inspection.results)),
    Inspection.Date = suppressWarnings(lubridate::mdy(as.character(Inspection.Date)))
  )
```

```{r load-closed}
closed <- read.csv("Closed_Food_Establishments.csv", stringsAsFactors = FALSE) %>%
  distinct() %>%
  mutate(
    Inspection.Results = str_trim(as.character(Inspection.Results)),
    Inspection.Date = suppressWarnings(lubridate::mdy(as.character(Inspection.Date)))
  )
```

## Quick data check

```{r}
closure_share <- nrow(closed) / nrow(food)

kable(
  tibble(
    Metric = c(
      "Inspection rows (deduplicated)",
      "Closed-establishment rows (deduplicated)",
      "Distinct closed establishments",
      "Date range in closure file",
      "Closure share (closed rows / all inspection rows)"
    ),
    Value = c(
      nrow(food),
      nrow(closed),
      n_distinct(closed$Establishment.ID),
      paste(
        format(min(closed$Inspection.Date, na.rm = TRUE), "%Y-%m-%d"),
        "to",
        format(max(closed$Inspection.Date, na.rm = TRUE), "%Y-%m-%d")
      ),
      sprintf("%.2f%%", 100 * closure_share)
    )
  )
)
```

## Visual 1: Overall inspection outcomes

```{r fig.width=8, fig.height=5}
food %>%
  count(Inspection.results, sort = TRUE) %>%
  mutate(Inspection.results = fct_reorder(Inspection.results, n)) %>%
  ggplot(aes(x = Inspection.results, y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = 1.05, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Inspection outcomes in FoodInspectionData",
    x = "Inspection result",
    y = "Count"
  ) +
  theme_minimal()
```

## Visual 2: Most common out-of-compliance issues among closures

```{r fig.width=9, fig.height=6}
violation_cols <- c(
  "Food.from.Approved.Source..C.",
  "Food.Protected.from.Contamination..C.",
  "Ill.Workers.Restricted..C.",
  "Proper.Hand.Washing..C.",
  "Cooling.Time.and.Temperature..C.",
  "Cold.Holding.Temperature..C.",
  "Hot.Holding.Temperature..C.",
  "Cooking.Time.and.Temperature..C.",
  "Reheating.Time.and.Temperature..C.",
  "Hot.and.Cold.Running.Water.Provided..C.",
  "Proper.Sewage.Disposal..C.",
  "Toxic.Substances...Pesticides",
  "Rodent.and.Insects"
)

closed %>%
  select(any_of(violation_cols)) %>%
  pivot_longer(everything(), names_to = "rule", values_to = "status") %>%
  mutate(status = str_trim(as.character(status))) %>%
  filter(!is.na(status), status == "Out of Compliance") %>%
  count(rule, sort = TRUE) %>%
  slice_head(n = 10) %>%
  mutate(rule = str_replace_all(rule, "\\.", " "),
         rule = fct_reorder(rule, n)) %>%
  ggplot(aes(x = rule, y = n)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Top out-of-compliance issues in closed establishments",
    x = "Inspection rule",
    y = "Count of out-of-compliance findings"
  ) +
  theme_minimal()
```

## Visual 3: Closure trend over time

```{r fig.width=9, fig.height=4}
closed %>%
  mutate(month = lubridate::floor_date(Inspection.Date, "month")) %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  ggplot(aes(month, n)) +
  geom_line(color = "darkgreen", linewidth = 0.9) +
  geom_point(color = "darkgreen", size = 1.8) +
  labs(title = "Monthly number of closure inspections", x = NULL, y = "Count") +
  theme_minimal()
```

## Additional view: Top cities with closure inspections

```{r fig.width=8, fig.height=5}
closed %>%
  mutate(City = str_to_title(str_trim(as.character(City)))) %>%
  filter(!is.na(City), City != "") %>%
  count(City, sort = TRUE) %>%
  slice_head(n = 10) %>%
  mutate(City = fct_reorder(City, n)) %>%
  ggplot(aes(x = City, y = n)) +
  geom_col(fill = "mediumpurple4") +
  coord_flip() +
  labs(
    title = "Top 10 cities by number of closure inspections",
    x = "City",
    y = "Closure inspection count"
  ) +
  theme_minimal()
```


## Note for your write-up

Suggested simple narrative:
- Most inspections are not closures, but closure events continue to happen each month.
- A small set of out-of-compliance issues appears repeatedly in closure cases.
- This helps target prevention efforts (training, follow-up, and temperature/sanitation checks).
