---
title: "DATA 205 Capstone"
subtitle: "Food inspections and closed establishments"
author: "Ousmane Kandji"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(tidyverse)
library(knitr)
library(lubridate)

# Cities need enough inspections for closure_rate to be stable.
MIN_INSPECTIONS_CITY <- 50L
```

## Project idea

This report uses:

- **FoodInspectionData.csv** for all inspection outcomes
- **Closed_Food_Establishments.csv** for inspections that ended in **Facility Closed**

Goal: provide a coherent workflow from raw data to cleaned data, EDA, and final visuals.

## 1) Load raw datasets

```{r load-raw}
food_raw <- read.csv("FoodInspectionData.csv", stringsAsFactors = FALSE)
closed_raw <- read.csv("Closed_Food_Establishments.csv", stringsAsFactors = FALSE)
```

## 2) Clean datasets

```{r clean-data}
clean_zip <- function(x) {
  digits <- str_replace_all(as.character(x), "[^0-9]", "")
  padded <- str_pad(str_sub(digits, 1, 5), width = 5, pad = "0")
  if_else(digits == "", NA_character_, padded)
}

food <- food_raw %>%
  distinct() %>%
  mutate(
    Inspection.results = str_trim(as.character(Inspection.results)),
    City = str_to_title(str_trim(as.character(City))),
    Zip = clean_zip(Zip),
    Inspection.Date = suppressWarnings(mdy(as.character(Inspection.Date)))
  )

closed <- closed_raw %>%
  distinct() %>%
  mutate(
    Inspection.Results = str_trim(as.character(Inspection.Results)),
    City = str_to_title(str_trim(as.character(City))),
    Zip = clean_zip(Zip),
    Inspection.Date = suppressWarnings(mdy(as.character(Inspection.Date)))
  )
```

## 3) Export cleaned datasets

```{r export-cleaned}
write_csv(food, "FoodInspectionData_cleaned.csv")
write_csv(closed, "Closed_Food_Establishments_cleaned.csv")
```

## 4) EDA summary

```{r eda-summary}
closure_row_share <- nrow(closed) / nrow(food)

kable(
  tibble(
    Metric = c(
      "Raw inspection rows",
      "Raw closure rows",
      "Inspection rows (deduplicated)",
      "Closed-establishment rows (deduplicated)",
      "Distinct closed establishments",
      "Date range in closure file",
      "Closure rows / all inspection rows (descriptive row ratio)"
    ),
    Value = c(
      nrow(food_raw),
      nrow(closed_raw),
      nrow(food),
      nrow(closed),
      n_distinct(closed$Establishment.ID, na.rm = TRUE),
      paste(
        format(min(closed$Inspection.Date, na.rm = TRUE), "%Y-%m-%d"),
        "to",
        format(max(closed$Inspection.Date, na.rm = TRUE), "%Y-%m-%d")
      ),
      sprintf("%.2f%%", 100 * closure_row_share)
    )
  )
)
```

## 5) EDA: Missingness check (selected fields)

```{r eda-missingness}
missing_tbl <- bind_rows(
  tibble(
    dataset = "food",
    variable = c("Inspection.results", "Inspection.Date", "Zip"),
    missing_n = c(
      sum(is.na(food$Inspection.results) | food$Inspection.results == ""),
      sum(is.na(food$Inspection.Date)),
      sum(is.na(food$Zip) | food$Zip == "")
    )
  ),
  tibble(
    dataset = "closed",
    variable = c("Inspection.Results", "Inspection.Date", "Zip"),
    missing_n = c(
      sum(is.na(closed$Inspection.Results) | closed$Inspection.Results == ""),
      sum(is.na(closed$Inspection.Date)),
      sum(is.na(closed$Zip) | closed$Zip == "")
    )
  )
)

kable(missing_tbl)
```

## Visual 1: Overall inspection outcomes

```{r visual-outcomes, fig.width=8, fig.height=5}
food %>%
  count(Inspection.results, sort = TRUE) %>%
  mutate(Inspection.results = fct_reorder(Inspection.results, n)) %>%
  ggplot(aes(x = Inspection.results, y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = 1.05, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Inspection outcomes in FoodInspectionData",
    x = "Inspection result",
    y = "Count"
  ) +
  theme_minimal()
```

## Visual 2: Most common out-of-compliance issues among closures

```{r visual-violations, fig.width=9, fig.height=6}
violation_cols <- c(
  "Food.from.Approved.Source..C.",
  "Food.Protected.from.Contamination..C.",
  "Ill.Workers.Restricted..C.",
  "Proper.Hand.Washing..C.",
  "Cooling.Time.and.Temperature..C.",
  "Cold.Holding.Temperature..C.",
  "Hot.Holding.Temperature..C.",
  "Cooking.Time.and.Temperature..C.",
  "Reheating.Time.and.Temperature..C.",
  "Hot.and.Cold.Running.Water.Provided..C.",
  "Proper.Sewage.Disposal..C.",
  "Toxic.Substances...Pesticides",
  "Rodent.and.Insects"
)

closed %>%
  select(any_of(violation_cols)) %>%
  pivot_longer(everything(), names_to = "rule", values_to = "status") %>%
  mutate(status = str_trim(as.character(status))) %>%
  filter(!is.na(status), status == "Out of Compliance") %>%
  count(rule, sort = TRUE) %>%
  slice_head(n = 10) %>%
  mutate(
    rule = str_replace_all(rule, "\\.", " "),
    rule = fct_reorder(rule, n)
  ) %>%
  ggplot(aes(x = rule, y = n)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Top out-of-compliance issues in closed establishments",
    x = "Inspection rule",
    y = "Count of out-of-compliance findings"
  ) +
  theme_minimal()
```

## Visual 3: Closure trend over time

```{r visual-trend, fig.width=9, fig.height=4}
closed %>%
  mutate(month = floor_date(Inspection.Date, "month")) %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  ggplot(aes(month, n)) +
  geom_line(color = "darkgreen", linewidth = 0.9) +
  geom_point(color = "darkgreen", size = 1.8) +
  labs(title = "Monthly number of closure inspections", x = NULL, y = "Count") +
  theme_minimal()
```

Closure counts by month are descriptive; any formal trend test assumes independent months and is optional exploratory follow-up.

## Top cities by closure inspection count

```{r visual-cities-count, fig.width=8, fig.height=5}
closed %>%
  filter(!is.na(City), City != "") %>%
  count(City, sort = TRUE) %>%
  slice_head(n = 10) %>%
  mutate(City = fct_reorder(City, n)) %>%
  ggplot(aes(x = City, y = n)) +
  geom_col(fill = "mediumpurple4") +
  coord_flip() +
  labs(
    title = "Top 10 cities by number of closure inspections",
    x = "City",
    y = "Closure inspection count"
  ) +
  theme_minimal()
```

## City closure rate (share of inspection rows)

`closure_rate = closure_rows / inspection_rows` per city. Cities with fewer than `r MIN_INSPECTIONS_CITY` inspections in the main file are excluded so tiny denominators do not dominate.

```{r city-risk-table}
city_risk <- food %>%
  filter(!is.na(City), City != "") %>%
  count(City, name = "inspection_rows") %>%
  left_join(
    closed %>%
      filter(!is.na(City), City != "") %>%
      count(City, name = "closure_rows"),
    by = "City"
  ) %>%
  mutate(
    closure_rows = replace_na(as.integer(closure_rows), 0L),
    closure_rate = closure_rows / inspection_rows
  ) %>%
  filter(inspection_rows >= MIN_INSPECTIONS_CITY)

kable(
  city_risk %>%
    arrange(desc(closure_rate), desc(inspection_rows)) %>%
    mutate(closure_rate_pct = sprintf("%.3f%%", 100 * closure_rate)) %>%
    select(City, inspection_rows, closure_rows, closure_rate_pct) %>%
    slice_head(n = 25)
)
```

```{r city-risk-plot, fig.width=9, fig.height=6}
city_risk %>%
  arrange(desc(closure_rate)) %>%
  slice_head(n = 15) %>%
  mutate(City = fct_reorder(City, closure_rate)) %>%
  ggplot(aes(x = City, y = closure_rate)) +
  geom_col(fill = "darkslateblue", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(round(100 * x, 3), "%")) +
  labs(
    title = sprintf(
      "City closure rate: closure rows / inspection rows (min %d inspections)",
      MIN_INSPECTIONS_CITY
    ),
    x = NULL,
    y = "Closure rate"
  ) +
  theme_minimal()
```

Higher closure rate means closure events in the extract account for a larger share of that city’s rows in the main dataset—interpret cautiously (linkage and extracts are not a perfect census of “all closures”).

## Note for your write-up

Suggested simple narrative:

- Most inspections are not closures, but closure events continue to happen each month.
- A small set of out-of-compliance issues appears repeatedly in closure cases.
- This helps target prevention efforts (training, follow-up, and temperature/sanitation checks).

Limitations to mention:

- `Closed_Food_Establishments.csv` is event-focused and much smaller than full inspection records.
- Results are descriptive EDA, not causal evidence.
- Some values may be missing or inconsistently entered in source files.
